---
title: "Intro to data visualization with ggplot2"
output:
  html_document:
    df_print: paged
---

In this session, we will work with data visualization using ggplot2. Although there are many systems in R for generating graphs, ggplot2 is the most versatile and elegant. We will use examples from the [R Graphics Cookbook](https://r-graphics.org).

> ggplot2 implements the **grammar of graphics**, a coherent system for describing and building graphs. With ggplot2, you can do more faster by learning one system and applying it in many places. [R4DS-Data Visualization](https://r4ds.had.co.nz/data-visualisation.html)

## Resources on ggplot2

- [The R Graphics Cookbook](http://www.cookbook-r.com/Graphs) by Winston Chang
- [Plotting Anything with ggplot2](https://youtu.be/h29g21z0a68) by Thomas Lin Pedersen (webinar)
- [R4DS-Data Visualization](https://r4ds.had.co.nz/data-visualisation.html) by Hadley Wickham & Garrett Grolemund
- [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/) is a more advanced and in-depth assessment of the theoretical underpinnings of ggplot2
- [Colorbrewer](https://colorbrewer2.org/) provide good recommendation for color selection

## Load required libraries

```{r library}
library(ggplot2)
library(reshape2)# contains the tips data

#install.packages("reshape2")
```

## Bar graphs of values

```{r}
dat <- data.frame(
  time = factor(c("Lunch","Dinner"), levels=c("Lunch","Dinner")),
  total_bill = c(14.89, 17.23)
)
dat
```

```{r}
# Very basic bar graph
ggplot(data=dat, aes(x=time, y=total_bill)) +
    geom_bar(stat="identity")
```

```{r}
# Map the time of day to different fill colors
ggplot(data=dat, aes(x=time, y=total_bill, fill=time)) +
    geom_bar(stat="identity")

## This would have the same result as above
# ggplot(data=dat, aes(x=time, y=total_bill)) +
#    geom_bar(aes(fill=time), stat="identity")
```

```{r}
# Add a black outline
ggplot(data=dat, aes(x=time, y=total_bill, fill=time)) +
    geom_bar(colour="#E69F00", stat="identity")

```

```{r}
# No legend, since the information is redundant
plt <- ggplot(data=dat, aes(x=time, y=total_bill, fill=time)) 

plt + geom_bar(colour="black", stat="identity") + guides(fill="none")
plt + geom_bar(colour="green", stat="identity") + guides(fill="none")


```

```{r}
?tips # get information about the tips dataset

head(tips)

# Bar graph of counts (more than 2)
ggplot(data=tips, aes(x=day)) +
    geom_bar(stat="count")
## Equivalent to this, since stat="bin" is the default:
# ggplot(data=tips, aes(x=day)) +
#    geom_bar()
```

```{r}

# Graphs with more variables
dat1 <- data.frame(
    sex = factor(c("Female","Female","Male","Male")),
    time = factor(c("Lunch","Dinner","Lunch","Dinner"), levels=c("Lunch","Dinner")),
    total_bill = c(13.53, 16.81, 16.24, 17.42)
)
dat1

ggplot(dat1, aes(total_bill)) + stat_ecdf(geom = "point")

```

```{r}
# Stacked bar graph -- this is probably not what you want
ggplot(data=dat1, aes(x=time, y=total_bill, fill=sex)) +
    geom_bar(stat="identity")
```

```{r}
# Bar graph, time on x-axis, color fill grouped by sex -- use position_dodge()
ggplot(data=dat1, aes(x=time, y=total_bill, fill=sex)) +
    geom_bar(stat="identity", position=position_dodge())

ggplot(data=dat1, aes(x=time, y=total_bill, fill=sex)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black")
```

```{r}
# Change colors
ggplot(data=dat1, aes(x=time, y=total_bill, fill=sex)) +
    geom_bar(stat="identity", position=position_dodge(), colour="black") +
    scale_fill_manual(values=c("#999999", "#E69F00"))
```

## Distribution plots

### Histogram - Single Group

```{r}
set.seed(1234)
dat2 <- data.frame(cond = factor(rep(c("A","B"), each=200)), 
                   rating = c(rnorm(200),rnorm(200, mean=.8)))
# View first few rows
head(dat2)
```

```{r}
## Basic histogram from the vector "rating". Each bin is .5 wide.
## These both result in the same output:
ggplot(dat2, aes(x=rating)) + geom_histogram(binwidth = 0.5)
# qplot(dat$rating, binwidth=.5)
```

```{r}
# Draw with black outline, white fill
ggplot(dat2, aes(x=rating)) +
    geom_histogram(binwidth=.5, colour="white", fill="black")
```

```{r}
# Density curve
ggplot(dat2, aes(x=rating)) + geom_density()
```

```{r}
# Histogram overlaid with kernel density curve
ggplot(dat2, aes(x=rating)) + 
    geom_histogram(aes(y=..density..),  # Histogram with density instead of count on y-axis
                   binwidth=.5,
                   colour="black", fill="white") +
    geom_density(alpha=0.2, fill="#FF6666")  # Overlay with transparent density plot
```

### Histogram - Multiple Groups

```{r}
# Overlaid histograms
ggplot(dat2, aes(x=rating, fill=cond)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
```

```{r}
# Interleaved histograms
ggplot(dat2, aes(x=rating, fill=cond)) +
    geom_histogram(binwidth=.5, position="dodge")
```

```{r}
# Density plots
ggplot(dat2, aes(x=rating, colour=cond)) + geom_density()
```

```{r}
# Density plots with semi-transparent fill
ggplot(dat2, aes(x=rating, fill=cond)) + geom_density(alpha=.3)

# stat_ecdf : empirical cumulative distribution function
ggplot(dat2, aes(rating)) + stat_ecdf(geom = "point")
ggplot(dat2, aes(rating)) + stat_ecdf(geom = "step")
ggplot(dat2, aes(rating)) + stat_ecdf(geom = "smooth")

# Multiple ECDFs based on condition
ggplot(dat2, aes(rating, color = cond)) + stat_ecdf(geom = "point")
```

### Box plots

```{r}
# A basic box plot
ggplot(dat2, aes(x=cond, y=rating)) + geom_boxplot()
```

```{r}
# A basic box with the conditions colored
ggplot(dat2, aes(x=cond, y=rating, fill=cond)) + geom_boxplot()
```

```{r}
# The above adds a redundant legend. With the legend removed:
ggplot(dat2, aes(x=cond, y=rating, fill=cond)) + geom_boxplot() +
    guides(fill="none")
```

```{r}
# With flipped axes
ggplot(dat2, aes(x=cond, y=rating, fill=cond)) + geom_boxplot() + 
    guides(fill="none") + coord_flip()
```

### Scatterplots

```{r}
set.seed(955)
# Make some noisily increasing data
dat3 <- data.frame(cond = rep(c("A", "B"), each=10),
                  xvar = 1:20 + rnorm(20,sd=3),
                  yvar = 1:20 + rnorm(20,sd=3))
head(dat3)

```

```{r}
ggplot(dat3, aes(x=xvar, y=yvar)) +
    geom_point(shape=5)      # Use hollow circles
```

```{r}
ggplot(dat3, aes(x=xvar, y=yvar)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth(method=lm)   # Add linear regression line 
                             #  (by default includes 95% confidence region)
```

```{r}
ggplot(dat3, aes(x=xvar, y=yvar)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth(method=lm,   # Add linear regression line
                se=FALSE)    # Don't add shaded confidence region
```

```{r}
ggplot(dat3, aes(x=xvar, y=yvar)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth()            # Add a loess smoothed fit curve with confidence region
#> `geom_smooth()` using method = 'loess'
```

## Facets

Split up your data by one or more variables and plot the subsets of data together.

```{r}
head(tips)

sp <- ggplot(tips, aes(x=total_bill, y=tip/total_bill)) + geom_point(shape=1)
sp
```

```{r}
# Divide by levels of "sex", in the vertical direction
sp + facet_grid(sex ~ .)
```

```{r}
# Divide by levels of "sex", in the horizontal direction
sp + facet_grid(. ~ sex)
```

```{r}
# Divide with "sex" vertical, "day" horizontal
sp + facet_grid(sex ~ day)
```

```{r}
# Divide by day, going horizontally and wrapping with 2 columns
sp + facet_wrap( ~ day, ncol=2)
```

## Saving plots

There are several commands which will direct output to a file instead of the screen. You must use the dev.off() command to tell R that you are finished plotting; otherwise your graph will not show up.

```{r saving-plots, eval=FALSE, include=TRUE}

# PDFs
pdf("plots.pdf") #svg
plot(...)
plot(...)
dev.off()

# PNG/TIFF
png("plot.png") # or tiff("plot.tiff")
plot(...)
dev.off()

# resizing plot
png("plot.png", width=480, height=240, res=120)
plot(...)
dev.off()

## ggplot2

# If you make plots with ggplot2 in a script or function, you must use the print() command to make the graphs actually get rendered.

ggsave("plot.pdf")

ggsave("plot.pdf", width=4, height=4)

# This will save a 400x400 file at 100 ppi
ggsave("plot.png", width=4, height=4, dpi=100)
```