---
title: "Intro to R and RStudio"
output:
  html_document:
    df_print: paged
---

## Background

R is a freely available language and environment for statistical computing and graphics which provides a wide variety of statistical and graphical techniques.

Free Resources

  - R for Data Science (https://r4ds.had.co.nz/index.html)
  - The Comprehensive R Archive Network (CRAN) (https://cloud.r-project.org)

## Basic R Syntax and Usage

---

### R as a calculator
```{r calculator}
1 + 100      # simple addition
1 +          # incomplete command. Press Control + C to cancel a command
3 + 5 * 2    # order of operations
3^2          # square 
```


### Mathematical functions
```{r math-functions}
sin(30)    # trigonometry functions
log(1)     # natural log
log10(10)  # log base-10
sqrt(4)    # square root
```


### Comparing things
```{r comparisons}
1 == 1    # equality (two equals sign)
1 != 2    # inequality
1 < 2     # less than
1 > 2     # greater than
1 >= 2    # greater than or equal to
```

### Vectors
```{r vectors}
1:5          # vector containing numbers 1 to 5
2^(1:5)      # vector containing 2 to the power of (1 to 5)
x <- 1:5     # assigning vector to a variable
2^x          # calculations on variables
```

## Variables

---

Variable names can contain letters, numbers, underscores and periods but no spaces. They must start with a letter or a period followed by a letter (they cannot start with a number nor an underscore). Variables beginning with a period are hidden variables. Different people use different conventions for long variable names, these include

- periods.between.words (avoid this as some functions use periods as well)
- underscores_between_words
- camelCaseToSeparateWords

What you use is up to you, but be **consistent.**

### Assigning values to variables
```
x <- 4     # assigns value to the variable x
x          # view output of x
```
output:
```
[1] 4
```
```
log(x)     # natural log of variable x
```
output:
```
[1] 1.386294
```
### Reassigning values to variables
```
x <- 100   # reassign value to x
x <- x+1
x
```
output:
```
[1] 101
```
```
y <- x^2  
```
output:
```
[1] 10201
```

### Variables can also store vectors, dataframes, and matrices
```
x <- 1:5     # assigning vector to a variable
2^x          # calculations on variables
```
output:
```
[1] 2 4 8 16 32
```

## Installing packages

---

In addition to base packages included with R, you can install additional packages that add novel functionalities to R (e.g., DESeq2 for differential expression analysis)

### Installing packages in CRAN
```{r packages, eval=FALSE}
install.packages()                # list all install packages 
install.packages("packagename")   # installing packagename
update.packages()                 # update all packages
update.packages("packagename")    # update packagename
library(packagename)              # make packagename available for use
```
### Optional: Installing packages in development through GitHub
```{r github-packages, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("Author/Packagename")
```

## Environment management

---

Managing environment in R
```{r environment-management}
ls()                # list all variables
rm(x)               # remove variable x
rm(list = ls())     # remove all variables
sessionInfo()       # print R version, OS, and loaded packages
```
Managing environment in RStudio (show demo)

## Example Exercise

---

What will the value of each variable be after each statement?
```{r example1, eval=FALSE}
mass <- 50
mass <- mass * 2
mass
```
output:
```
[1]
```
```{r}
age <- 103
age <- age -50
age
```
output:
```
[1]
```

Write a code comparing mass to age
```{r example2, eval=FALSE}
age ? mass    # is age not equal to mass?
mass ? age     # is mass greater than age?
```


Clean up the environment by deleting the mass and age variables
```{r example3, eval=FALSE}
rm(?) # removes all variables
rm(?) # removes age variable
rm(?) # removes mass variable
rm(?) # removes mass and age variable
```
